---
title: "Untitled"
author: "Charlene Garridos"
date: '2023-04-03'
output: 
  pdf_document:
    highlight: tango
---

```{r}
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
```

```{r}
salespeople <- read.csv(url)
local_salespeople <- read.csv("salespeople.csv")
```

# create df_report function
```{r, echo=TRUE}
df_report <- function(df){paste("This dataframe contains", nrow(df), "rows and", ncol(df), "columns. There are", sum(is.na(df)), "NA entries")}
```

```{r}
df_report(mtcars)
```

```{r}
paste("This is how the paste", "functioning works")
```
# installing packages
```{r, error=TRUE}
my_packages <- c("MASS", "DescTools", "dplyr")
install.packages(my_packages)
```

```{r}
installed.packages("MASS")
```
 
```{r}
my_packages <- c("DescTools", "dplyr")
```

```{r}
installed.packages("my_packages")
```

```{r}
library(MASS)
```

```{r}
help(package = "MASS")
```
# The pipe operator
```{r}
sales <- subset(salespeople, subset = sales < 500)
```

```{r}
mean(sales$sales)
```

```{r}
mean(subset(salespeople, subset = sales <500) $ sales)
```

```{r}
mean(subset(salespeople$sales, subset = salespeople$sales<500))
```

# Load magrittr library to get the pipe operator
```{r}
library(magrittr)
```
#Use the pipe operator to lay out the steps more logically
```{r}
subset(salespeople$sales, subset = salespeople$sales < 500) %>%
mean()
```
```{r}
library(magrittr)
subset(salespeople, subset = sales < 500)$sales %>%
mean()
```

```{r}
salespeople$sales %>% # start with all data
  subset(subset = salespeople$sales < 500) %>% # get the subsetted data
  mean() %>% # take the mean value
  round() # round to the nearest time
```
# Errors, warnings and messages
```{r, error=TRUE}
subset(salespeople, subset = sales = 700)
```
```{r}
subset(salespeople, subset = sales == 700)
```
 
```{r, error=TRUE}
head[salespeople]
```
 
```{r}
head(salespeople)
```
 
```{r}
salespeople[1,0]
```

# plotting and graphing
## scatter plot of customer_rate against sales
```{r}
plot(x =  salespeople$sales, y =  salespeople$customer_rate, xlab = "Sales ($m)", ylab = "Average customer rating", main = "Scatterplot of Slaes vs Customer rating")
```

# Convert performance ratings back to numeric data type for histogram
```{r}
salespeople$performance <- as.numeric(salespeople$performance)
```
#histogram of performance ratings
```{r}
hist(salespeople$performance)
```
```{r}
hist(salespeople$performance, xlab = "Performance Ratings", ylab = "none", main = "Histogram of performance ratings")
hist(salespeople$performance, breaks = 0:4)
```

## box plot of sales by performance rating
```{r}
boxplot(formula = sales ~ performance, data = salespeople,
        xlab = "Performance Rating", ylab = "Sales ($m)",
        main = "Boxplot of Sales by Performance Rating")
```

# Specialist plotting and graphing packages
## convert performance and promotion to categorical

```{r}
install.packages(
   "ggplot2",
   repos = c("http://rstudio.org/_packages",
   "http://cran.rstudio.com")
)
```

```{r}
library(GGally)
```

```{r}
salespeople$promoted <- as.factor(salespeople$promoted)
salespeople$performance <- as.factor(salespeople$performance)
```

## Pairplot of salespeople
```{r, warning=FALSE}
GGally::ggpairs(salespeople)
```


 
 
 
 
 
